---
title: "class19"
author: "Steven Gan"
format: pdf
toc: true
date: 2022-12-03
theme:
  light: faltly
  dark: darkly
---

# 1. Investigating pertussis cases by year

Q1:

```{r}
# install.packages("datapasta")
library(datapasta)

cdc <- data.frame(
                                 Year = c(1922L,1923L,1924L,1925L,
                                          1926L,1927L,1928L,1929L,1930L,1931L,
                                          1932L,1933L,1934L,1935L,1936L,
                                          1937L,1938L,1939L,1940L,1941L,1942L,
                                          1943L,1944L,1945L,1946L,1947L,
                                          1948L,1949L,1950L,1951L,1952L,
                                          1953L,1954L,1955L,1956L,1957L,1958L,
                                          1959L,1960L,1961L,1962L,1963L,
                                          1964L,1965L,1966L,1967L,1968L,1969L,
                                          1970L,1971L,1972L,1973L,1974L,
                                          1975L,1976L,1977L,1978L,1979L,1980L,
                                          1981L,1982L,1983L,1984L,1985L,
                                          1986L,1987L,1988L,1989L,1990L,
                                          1991L,1992L,1993L,1994L,1995L,1996L,
                                          1997L,1998L,1999L,2000L,2001L,
                                          2002L,2003L,2004L,2005L,2006L,2007L,
                                          2008L,2009L,2010L,2011L,2012L,
                                          2013L,2014L,2015L,2016L,2017L,2018L,
                                          2019L),
         No..Reported.Pertussis.Cases = c(107473,164191,165418,152003,
                                          202210,181411,161799,197371,
                                          166914,172559,215343,179135,265269,
                                          180518,147237,214652,227319,103188,
                                          183866,222202,191383,191890,109873,
                                          133792,109860,156517,74715,69479,
                                          120718,68687,45030,37129,60886,
                                          62786,31732,28295,32148,40005,
                                          14809,11468,17749,17135,13005,6799,
                                          7717,9718,4810,3285,4249,3036,
                                          3287,1759,2402,1738,1010,2177,2063,
                                          1623,1730,1248,1895,2463,2276,
                                          3589,4195,2823,3450,4157,4570,
                                          2719,4083,6586,4617,5137,7796,6564,
                                          7405,7298,7867,7580,9771,11647,
                                          25827,25616,15632,10454,13278,
                                          16858,27550,18719,48277,28639,32971,
                                          20762,17972,18975,15609,18617)
       )
```

```{r}
library(ggplot2)

p <- ggplot(cdc) +
  aes(Year, No..Reported.Pertussis.Cases) +
  geom_point() +
  geom_line() +
  labs(x = "Year", y = "Number of cases",
       title = "Pertussis Cases by Year (1922-2019)")
p
```


# 2. A tale of two vaccines (wP & aP)

Q2: 

```{r}
p + geom_vline(xintercept = 1946, linetype = "dashed", col = "blue") +
  geom_vline(xintercept = 1996, linetype = "dashed", col = "red") +
  annotate(geom = "text", x = 1950, y = 250000, label = "wP", 
           col = "blue", size = 5) +
  annotate(geom = "text", x = 2000, y = 250000, label = "aP", 
           col = "red", size = 5)
```

After the application of wP, number of cases dramatically decreased to approaches 0, indicating the robust effect of vaccination. After the change of wP to aP, number of cases slightly increased, but still significantly lower than without vaccination.

Q3:

Number of cases increased slightly after introducing aP. A possible explanation is the effect of aP is weaker than wP, since aP is made from acellular antigen with weakened toxin.


# 3. Exploring CMI-PB data

## The CMI-PB API returns JSON data

```{r}
# install.packages("jsonlite")
library(jsonlite)
```

```{r}
subject <- read_json("https://www.cmi-pb.org/api/subject", 
                     simplifyVector = TRUE) 

head(subject)
```

Q4:

```{r}
table(subject$infancy_vac)
```

Q5:

```{r}
table(subject$biological_sex)
```

Q6:

```{r}
table(subject$biological_sex, subject$race)
```

## Side-Note: Working with dates

```{r, message = FALSE}
library(lubridate)

today()
today() - ymd("2001-02-11")
time_length(today() - ymd("2005-02-27"), "years")
```

Q7:

(i)

```{r, message = FALSE}
library(dplyr)

subject$age <- time_length(today() - ymd(subject$year_of_birth), "years")

subject %>% 
  filter(infancy_vac == "wP") %>%
  summarise(Mean = mean(age)) %>%
  round()
```

(ii)

```{r}
subject %>% 
  filter(infancy_vac == "aP") %>%
  summarise(Mean = mean(age)) %>%
  round()
```

(iii)

```{r}
t.test(filter(subject, infancy_vac == "wP")$age, 
       filter(subject, infancy_vac == "aP")$age)
```

Q8:

```{r}
time_length(ymd(subject$date_of_boost) - ymd(subject$year_of_birth), "years")
```

Q9:

```{r}
ggplot(subject) +
  aes(time_length(age, "year"),
      fill = as.factor(infancy_vac)) +
  geom_histogram(show.legend = FALSE) +
  facet_wrap(vars(infancy_vac), nrow = 2) 
```

They are significantly different

## Joining multiple tables

```{r}
specimen <- read_json("https://www.cmi-pb.org/api/specimen", 
                      simplifyVector = TRUE) 
titer <- read_json("https://www.cmi-pb.org/api/ab_titer", 
                   simplifyVector = TRUE) 
```

Q9:

```{r}
meta <- inner_join(specimen, subject)
dim(meta)
```

Q10:

```{r}
abdata <- inner_join(titer, meta)
dim(abdata)
```

Q11:

```{r}
table(abdata$isotype)
```

Q12:

```{r}
table(abdata$visit)
```

The number of visit 8 specimens are significantly lower than others.


# 4. Examine IgG1 Ab titer levels

```{r}
ig1 <- abdata %>%
  filter(isotype == "IgG1" & visit != 8)
```

Q13:

```{r}
ggplot(ig1) +
  aes(MFI, antigen) +
  geom_boxplot() + 
  facet_wrap(vars(visit), nrow = 2)
```

Q14:

FIM2/3 exhibited most significant differences in the level of IgG1 antibody titers overtime.

```{r}
ggplot(ig1) +
  aes(MFI, antigen, col = infancy_vac ) +
  geom_boxplot(show.legend = FALSE) + 
  facet_wrap(vars(visit), nrow = 2) +
  theme_bw()
```

```{r}
ggplot(ig1) +
  aes(MFI, antigen, col = infancy_vac ) +
  geom_boxplot(show.legend = FALSE) + 
  facet_wrap(vars(infancy_vac, visit), nrow = 2)
```

Q15:

```{r}
filter(ig1, antigen == "Measles") %>%
  ggplot() +
  aes(MFI, col = infancy_vac) +
  geom_boxplot(show.legend = F) +
  facet_wrap(vars(visit)) +
  theme_bw()
```

```{r}
filter(ig1, antigen == "FIM2/3") %>%
  ggplot() +
  aes(MFI, col = infancy_vac) +
  geom_boxplot(show.legend = F) +
  facet_wrap(vars(visit)) +
  theme_bw()
```

```{r}
filter(ig1, antigen == "FIM2/3" | antigen == "Measles") %>%
  ggplot() +
  aes(MFI, antigen, col = infancy_vac) +
  geom_boxplot(show.legend = F) +
  facet_wrap(vars(visit)) +
  theme_bw()
```


```{r}
filter(ig1, antigen == "FIM2/3" | antigen == "Measles") %>%
  ggplot() +
  aes(visit, MFI, col = infancy_vac, antigen, group = subject_id) +
  geom_point(show.legend = F) +
  geom_line(alpha = 0.2) +
  facet_wrap(vars(antigen)) +
  theme_bw()

filter(ig1, antigen == "FIM2/3" | antigen == "Measles") %>%
  ggplot() +
  geom_point(aes(visit, MFI, col = infancy_vac), show.legend = F) +
  geom_line(aes(visit, MFI, group = subject_id), alpha = 0.1) + 
  geom_smooth(aes(visit, MFI, col = infancy_vac), show.legend = F) +
  facet_wrap(vars(antigen)) +
  theme_bw()
```

Q16:

FIM2/3 rises and peaks at visit 5, whereas Measles do not rise at all. 

Q17: 

```{r}
t.test(filter(ig1, antigen == "FIM2/3" & infancy_vac == "wP" & visit == 5)$MFI,
       filter(ig1, antigen == "FIM2/3" & infancy_vac == "aP" & visit == 5)$MFI)
```

There are no significant difference between wP and aP responses.


# 5. Obtaining CMI-PB RNASeq data

```{r}
url <- "https://www.cmi-pb.org/api/v2/rnaseq?versioned_ensembl_gene_id=eq.ENSG00000211896.7"

rna <- read_json(url, simplifyVector = TRUE) 
```

```{r}
ssrna <- inner_join(rna, meta)
```

Q18:

```{r}
ggplot(ssrna) +
  aes(visit, tpm, group = subject_id) +
  geom_point() +
  geom_line(alpha = 0.2)
```

Q19:

The expression of IGHG1 gene peaks at visit 4, and drops back to basal line quickly after it.

Q20: 

The peaking of transcription level is a visit early than the peaking of antibody titer data. This could be explained by the fact that the RNA expression changes is prior to protein level changes.

```{r}
ggplot(ssrna) +
  aes(tpm, col = infancy_vac) +
  geom_boxplot() +
  facet_wrap(vars(visit))
```

```{r}
ssrna %>%  
  filter(visit == 4) %>% 
  ggplot() +
    aes(tpm, col=infancy_vac) +
    geom_density() +
    geom_rug()
```


# 6. Working with larger datasets

```{r}
rnaseq <- read.csv("2020LD_rnaseq.csv")

head(rnaseq,3)
dim(rnaseq)
```

# Working with long format data

```{r}
n_genes <- table(rnaseq$specimen_id)
head(n_genes , 10)
```

```{r}
all(n_genes[1] == n_genes)
```

## Convert to “wide” format

```{r}
library(tidyr)

rna_wide <- rnaseq %>% 
  select(versioned_ensembl_gene_id, specimen_id, tpm) %>%
  pivot_wider(names_from = specimen_id, values_from = tpm)

dim(rna_wide)
```

## Filtering

```{r}
rna_wide <- as.data.frame(rna_wide)
rownames(rna_wide) <- rna_wide[,1]
rna_wide <- rna_wide[-1]

rna <- rna_wide[rowSums(rna_wide == 0) == 0,]
```

## PCA

```{r}
rna.pr <- prcomp(t(rna), scale = T)
summary(rna.pr)
```

```{r}
library(ggrepel)

rna.pr$x %>%
  as.data.frame() %>%
  ggplot() +
  aes(PC1, PC2, label = rownames(rna.pr$x)) +
  geom_point() +
  geom_text_repel(size = 3.3, max.overlaps = 5)
```

```{r, message = FALSE}
library(factoextra)

fviz_eig(rna.pr, addlabels = TRUE)
```

```{r}
rna.norm <- t(scale(t(rna)))
heatmap(t(rna.norm))
heatmap(rna.pr$x)
```