---
title: "class10"
author: "Steven Gan"
format: html
toc: true
date: 2022-10-28
theme:
  light: faltly
  dark: darkly
---

# 1. Importing candy data

```{r}
candy <- read.csv(file = "https://raw.githubusercontent.com/fivethirtyeight/data/master/candy-power-ranking/candy-data.csv", 
                  row.names = 1)

head(candy)
```

Q1:
```{r}
nrow(candy)
```

Q2:
```{r}
sum(candy$fruity)
```

# 2. What is your favorate candy?

```{r}
candy["Twix", ]$winpercent
```

Q3:
```{r}
print("Milky Way")
candy["Milky Way", ]$winpercent
```

Q4:
```{r}
candy["Kit Kat", ]$winpercent
```

Q5:
```{r}
candy["Tootsie Roll Snack Bars", ]$winpercent
```

```{r}
# install.packages("skimr")
library(skimr)

skim(candy)
```

Q6:
The variable "winpercent" is different in scale, which is from 0 to 100.

Q7:
1 and 0 means whether that type of candy contains chocolate or not.

Q8:
```{r}
library(ggplot2)

ggplot(candy, aes(x = winpercent)) +
  geom_histogram(bins = 20, fill = "yellow", col = "red")
```

Q9:
The distribution of winpercent values is not completely symmetrical.

Q10:
Below

Q11:
```{r}
chocoWinper <- candy[as.logical(candy$chocolate), ]$winpercent
fruWinper <- candy[as.logical(candy$fruity), ]$winpercent

if (mean(chocoWinper) > mean(fruWinper)) 
  {print("Higher")} else {print("Lowwer")}
```

Q12:
```{r}
t.test(chocoWinper, fruWinper)
```
Different statistically significant

# 3. Overall Candy Rankings

Q13:
```{r}
library(dplyr)

candy %>% 
  arrange(winpercent) %>%
  head(5) %>%
  row.names()
```

Q14:
```{r}
candy %>% 
  arrange(desc(winpercent)) %>%
  head(5) %>%
  row.names()
```

Q15:
```{r}
ggplot(candy) + 
  aes(winpercent, rownames(candy)) +
  geom_col(width = 0.7) + 
  theme(text = element_text(size = 5), element_line(size = 0.3), 
        aspect.ratio = 3)

# ggsave("mybarplot.png")
```

Q16:
```{r}
ggplot(candy) + 
  aes(winpercent, reorder(rownames(candy), winpercent)) +
  geom_col(width = 0.7) + 
  theme(text = element_text(size = 5), element_line(size = 0.3), 
        aspect.ratio = 3)
```

```{r}
my_cols = rep("black", nrow(candy))
my_cols[as.logical(candy$chocolate)] = "chocolate"
my_cols[as.logical(candy$bar)] = "brown"
my_cols[as.logical(candy$chocolate) & as.logical(candy$bar)] = "blue"
my_cols[as.logical(candy$fruity)] = "pink"
```

```{r}
ggplot(candy) + 
  aes(winpercent, reorder(rownames(candy),winpercent)) +
  geom_col(width = 0.7, fill = my_cols) + 
  theme(text = element_text(size = 5), element_line(size = 0.3), 
        aspect.ratio = 3)

# gsave("ChocoBarFru.png")
```

Q17:
Sixlets

Q18:
Starburst

# 4. Taking a look at pricepercent

```{r}
library(ggrepel)

# How about a plot of price vs win
ggplot(candy) +
  aes(winpercent, pricepercent, label = rownames(candy)) +
  geom_point(col = my_cols) + 
  geom_text_repel(col = my_cols, size = 3.3, max.overlaps = 5)

```

Q19:
Probably Sour Patch Kids, Starburst, ReeseÕs Miniatures

Q20:
```{r}
candy %>%
  arrange(desc(pricepercent)) %>%
  head(5) %>%
  row.names()

candy %>%
  arrange(desc(pricepercent)) %>%
  head(5) %>%
  arrange(winpercent) %>%
  head(1) %>%
  row.names()
```

Q21:
```{r}
ggplot(candy) + 
  aes(pricepercent, reorder(rownames(candy), pricepercent)) +
  geom_col(width = 0.7) + 
  theme(text = element_text(size = 5), element_line(size = 0.3), 
        aspect.ratio = 3)
```

```{r}
ggplot(candy) +
  aes(pricepercent, reorder(rownames(candy), pricepercent)) +
  geom_segment(aes(yend = reorder(rownames(candy), pricepercent), 
                   xend = 0), col = "gray40") +
  geom_point() +
  theme(text = element_text(size = 5), element_line(size = 0.3), 
        aspect.ratio = 3)
```

# 5 Exploring the correlation structure

```{r}
library(corrplot)

cij <- cor(candy)
corrplot(cij)
```

Q22:
Chocolate and fruity are anti-correlated

Q23:
Chocolate and Winpercent are correlated

# 6. Principal Component Analysis

```{r}
pca <- prcomp(candy, scale = T)
summary(pca)
```

```{r}
library(factoextra)
fviz_eig(pca, addlabels = TRUE)
```

```{r}
plot(pca$x[,1:2], col = my_cols, pch = 16)
```

```{r}
my_data <- cbind(candy, pca$x[,1:3])
```

```{r}
p <- ggplot(my_data) + 
        aes(x = PC1, y = PC2, 
            size = winpercent / 100,  
            text = rownames(my_data),
            label = rownames(my_data)) +
        geom_point(col = my_cols)

p
```

```{r}
p + geom_text_repel(size = 3.3, col = my_cols, max.overlaps = 5) + 
  theme(legend.position = "none") +
  labs(title = "Halloween Candy PCA Space",
       subtitle = "Colored by type: chocolate bar (dark brown), chocolate other (light brown), fruity (red), other (black)",
       caption = "Data from 538")
```

```{r}
# install.packages("plotly")
library(plotly)

ggplotly(p)
```

```{r}
par(mar = c(8,4,2,2))
barplot(pca$rotation[,1], las = 2, ylab = "PC1 Contribution")
```

Q24:
Fruity, hard, and pluribus are in positive direction.
Cutomarily it seems to make sense that fruity candies are hard candies packaged in bags or boxs.