---
title: "class13"
author: "Steven Gan"
format: html
toc: true
date: 2022-11-09
theme:
  light: faltly
  dark: darkly
---

# Section 1. Differential Expression Analysis

```{r, message = FALSE}
library(DESeq2)
```

```{r}
colData = read.csv(file = "GSE37704_metadata.csv", row.names = 1)
countData = read.csv(file = "GSE37704_featurecounts.csv", row.names = 1)
```

Q1:

```{r}
countData <- countData[,-1]
```

Q2:

```{r}
# Delete rows that have either all 0 in WT or KD
countData = countData[((rowSums(countData[,1:3] != 0) != 0) &
                       (rowSums(countData[,4:6] != 0) != 0)),]
nrow(countData) # 14861

# Deplete rows with all 0
# countData = countData[rowSums(countData != 0) != 0,]
# nrow(countData) #15975
```

## PCA analysis

```{r}
pca <- prcomp(t(countData), scale = TRUE)
summary(pca)
```

```{r}
plot(pca$x[,1], pca$x[,2], col = as.factor(colData$condition), pch = 16)
```



## Running DESeq2

```{r}
dds = DESeqDataSetFromMatrix(countData = countData,
                             colData = colData,
                             design = ~condition)
dds = DESeq(dds)
```

```{r}
dds
```

```{r}
res = results(dds)
```

Q3: 

```{r}
summary(res)
```

## Volcono plot

```{r}
plot(res$log2FoldChange, -log(res$padj))
```

Q4: 

```{r}
mycols <- rep("gray", nrow(res))

mycols[abs(res$log2FoldChange) > 2] <- "red"

inds <- (res$padj < 0.01) & (abs(res$log2FoldChange) > 2)
mycols[inds] <- "blue"

plot(res$log2FoldChange, -log(res$padj), col= mycols, 
     xlab = "Log2(FoldChange)", ylab = "-Log(P-value)" )
```

## Adding gene annotation

Q5: 

```{r}
library("AnnotationDbi")
library("org.Hs.eg.db")

res$symbol = mapIds(org.Hs.eg.db,
                    keys = row.names(res), 
                    keytype = "ENSEMBL",
                    column = "SYMBOL",
                    multiVals = "first")

res$entrez = mapIds(org.Hs.eg.db,
                    keys = row.names(res),
                    keytype = "ENSEMBL",
                    column = "ENTREZID",
                    multiVals = "first")

res$name =   mapIds(org.Hs.eg.db,
                    keys = row.names(res),
                    keytype = "ENSEMBL",
                    column = "GENENAME",
                    multiVals = "first")

head(res, 10)
```

Q6:

```{r}
res = res[order(res$pvalue),]
write.csv(res, file = "deseq_results.csv")
```


# Section 2. Pathway Analysis

```{r, message = FALSE}
# BiocManager::install(c("pathview", "gage", "gageData"))

library(pathview)
library(gage)
library(gageData)
```

```{r}
data(kegg.sets.hs)
data(sigmet.idx.hs)

kegg.sets.hs = kegg.sets.hs[sigmet.idx.hs]

head(kegg.sets.hs, 3)
```

```{r}
foldchanges = res$log2FoldChange
names(foldchanges) = res$entrez
head(foldchanges)
```

```{r}
keggres = gage(foldchanges, gsets = kegg.sets.hs)
```

```{r}
attributes(keggres)
```

```{r}
head(keggres$less)
```

```{r, message = FALSE}
pathview(gene.data = foldchanges, pathway.id = "hsa04110")

pathview(gene.data = foldchanges, pathway.id = "hsa04110", kegg.native = FALSE)
```

```{r}
keggrespathways <- rownames(keggres$greater)[1:5]

keggresids = substr(keggrespathways, start=1, stop=8)
keggresids
```

```{r, message = FALSE}
pathview(gene.data = foldchanges, pathway.id = keggresids, species = "hsa")
```

Q7:

```{r}
keggrespathways <- rownames(keggres$less)[1:5]

keggresids = substr(keggrespathways, start = 1, stop = 8)
keggresids
```

```{r, message = FALSE}
pathview(gene.data = foldchanges, pathway.id = keggresids, species = "hsa")
```

# Section 3. Gene Ontology (GO)

```{r}
data(go.sets.hs)
data(go.subs.hs)

gobpsets = go.sets.hs[go.subs.hs$BP]

gobpres = gage(foldchanges, gsets = gobpsets, same.dir = TRUE)

lapply(gobpres, head)
```

# Section 4. Reactome Analysis

```{r}
sig_genes <- res[res$padj <= 0.05 & !is.na(res$padj), "symbol"]
print(paste("Total number of significant genes:", length(sig_genes)))
```

```{r}
write.table(sig_genes, file = "significant_genes.txt", 
            row.names = FALSE, col.names = FALSE, quote = FALSE)
```

Q8: 

"Endosomal/Vacuolar pathway" exhibited the lowest entities p value, followed by "Cell cycle, mitotic".
However, in terms of entities found, the latter exhibits 410 entities, which is far more than the entities found for the former. Possible explanation for the difference results between the methods is that the reactome analysis only analysed based on the genes names with significant chances, however KEGG also considered the fold change level of genes.

# Section 5. GO online (OPTIONAL)

Q9:

The rank-1 pathway is "regulation of cell migration involved in sprouting angiogenesis", followed by "negative regulation of mitotic nuclear division". Certain relations could be argued in between this result and the result from KEGG & GO. Negative mitotic nuclear division appeared in both methods, and cell migration is related to homophile cell adhesion and cell cell adhesion. Similarly to reactome analysis, GO online also ignore the fold change level and only focus on what genes were regulated.


# GO ploting
```{r, message = FALSE}
# BiocManager::install("clusterProfiler")
library(clusterProfiler)
library(tidyverse)
```

```{r}
x <- as.data.frame(res)
y <- filter(x, log2FoldChange < -2 , padj < 0.05)
```

```{r}
ecc <- enrichGO(rownames(y),
                OrgDb = org.Hs.eg.db,
                keyType = "ENSEMBL",
                ont = "BP")
```

```{r}
barplot(ecc)
dotplot(ecc)
goplot(ecc)
```